---
title: "Dominance plot in 2 to N dimensions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,  eval = TRUE, echo = TRUE, message = FALSE}
library(dominatR)
library(airway)
library(dominatRData)
library(ggplot2)
```

# Data

The data used in this article comes from two different sources:

1.  A `SummarizedExperiment` object from the library `airway`
2.  A `data.frame` retrieved from the supplementary library `dominatRData`

The purpose of using two different sources is meant to show the capacity of dominatR to process different objects.

```{r}
### summarized experiment
data("airway")
se <-airway

#Let's normalize the se object and subset
set.seed(123)
se <- cpm_normalization(se, log_trans = FALSE, new_assay_name = 'cpm_norm')
idx <- sample(seq_len(nrow(se)), size = min(25000, nrow(se)))
se <- se[idx, ]
colnames(se) <- paste0('S', seq_len(8))

# dataframe
data("rnapol_score")
df1 = rnapol_score
df1 = df1[,5:7]

data("atac_tissue_score")
df2 = atac_tissue_score

```


# Visualization Functions

Visual representation is essential for interpreting the structure, dominance, and variability of features across samples or conditions.

Our package offers a collection of entropy-based visualization functions designed for different analytical perspectives:

-   `plot_rope()`\
    Compares two numeric vectors using a central “rope” layout to visualize dominance asymmetry and entropy filtering.

-   `plot_triangle()`\
    Visualizes three variables in a ternary layout, highlighting balance or dominance among triplets.

-   `plot_circle()`\
    Displays each sample's entropy and average magnitude in a polar coordinate layout.

-   `plot_circle_frequency()`\
    Summarizes the density of entropy-magnitude bins using circular heat segments.

Let’s now explore each visualization function with real data examples.

## plot_rope(): Rope Plot for Binary Feature Dominance

This function compares two numeric vectors (e.g., expression in Condition A vs. B) using a "rope-like" 1D dominance visualization. Each sample is classified by its relative dominance, optionally filtered by entropy or magnitude thresholds.

This function is ideal for:

-   Comparing two groups of measurements across matched samples or features.

## Using a Summarized Experiment

Using the se object and selecting two columns. By default the function considers all the possible entropy values and magnitude values.

### Basic Usage

```{r}
sample1 <- "S1"
sample2 <- "S2"
res_rope <- plot_rope(
    x = se,
    push_text =  1.2,
    col = c('#7eb1d5', '#fa8451'),
    column_name = c(sample1, sample2), assay_name = 'cpm_norm'
)
```
The points that locate at the center of the rope, are those shared across both variables. As the points start moving towards the end of the rope, feature dominance is appreciated by the respective variable

The following aesthetics can be modified:

-   `rope_widht`

-   `rope_color`

-   `rope_border`

-   `col`

-   `col_bg`

-   `pch`

-   `pch_bg`

-   `cex`

-   `title`

Now let's explore the features the plotting function offers.

### Example 1: Low entropy

Considering observations with low entropy (High Dominance) and any level of expression. Adding a title to the plot and changing the colors for the observations. The observations that fall outside of this range will be colored with `col_bg` which by default is `whitesmoke`

```{r}
res_rope = plot_rope(
    x = se,
    push_text = 1.2,
    column_name = c(sample1, sample2),
    col = c('#7eb1d5', '#fa8451'),
    entropyrange = c(0, 0.1),
    maxvaluerange = c(0.1, Inf),
    assay_name = 'cpm_norm'
)

title(main = "SE Input: Low Entropy", font.main = 1)
```

### Example 2: Medium entropy

```{r}
res_rope = plot_rope(
    x = se,
    push_text = 1.2,
    column_name = c(sample1, sample2),
    col = c('#7eb1d5', '#fa8451'),
    entropyrange = c(0.15, 0.8),
    maxvaluerange = c(0.1, Inf),
    assay_name = 'cpm_norm'
)

title(main = "SE Input: Medium Entropy", font.main = 1)
```

### Example 3: High entropy

```{r}
res_rope = plot_rope(
    x = se,
    push_text = 1.2,
    column_name = c(sample1, sample2),
    col = c('#7eb1d5', '#fa8451'),
    entropyrange = c(0.8, 1),
    maxvaluerange = c(0.1, Inf),
    assay_name = 'cpm_norm'
)


title(main = "SE Input: High Entropy", font.main = 1)

```

### Example 4: Retrieve output data

Data retrieval can be achieved by modifying the attribute `output_table` . The data contains information used for the plotting (This applies for any data structure used)

-   **`a`, `b`**\
    The original values from each of the two input columns used

-   **`comx`, `comy`**\
    The computed Cartesian coordinates for each point on the “rope”.

-   **`color`**\
    The fill color (as a string) actually used for that point.

-   **`entropy`**\
    The Shannon entropy score for that feature across all columns.

-   **`maxvalue`**\
    The mean (or maximum) expression value used to scale point size (or filter).

```{r}
res_rope = plot_rope(
    x = se,
    push_text = 1.2,
    column_name = c(sample1, sample2),
    output_table = TRUE,
    col = c('#7eb1d5', '#fa8451'),
    entropyrange = c(0.8, 1),
    maxvaluerange = c(0, Inf), 
    assay_name = 'cpm_norm'
)

head(res_rope)

```

## Using a matrix or data.frame input

### Basic Usage

```{r}

sample1 = 'pol1'
sample2 = 'pol2'


res_rope = plot_rope(
    x = df1,
    push_text = 1.2,
    col = c('#a9c4d6', '#ffc9b3'),
    column_name = c(sample1, sample2),
)

title(main = 'Basic Usage')
```

### Example 1. Low entropy filtering (0-0.1)

```{r}
res_rope = plot_rope(
    x = df1,
    column_name = c(sample1, sample2),
    push_text = 1.2,
    col = c('#a9c4d6', '#ffc9b3'),
    entropyrange = c(0, 0.1),
)

title(main = 'Low Entropy Filtering')
```

### Example 2. High entropy filtering (0.8-1)

```{r}
res_rope = plot_rope(
    x = df1,
    column_name = c(sample1, sample2),
    push_text = 1.2,
    col = c('#a9c4d6', '#ffc9b3'),
    entropyrange = c(0.8, 1),
)

title(main = 'High Entropy Filtering')
```
### Miscellaneous

Changing aesthetics attributes

```{r}
res_rope = plot_rope(
    x = df1,
    column_name = c(sample1, sample2),
    rope_color = 'white',
    col = c('#001052', '#fa4c07'),
    col_bg = 'gray',
    pch_bg = 2,
    pch = c(10, 18),
    entropyrange = c(0, 0.5),
)

title(main = 'Modifying Attributes')
```

## plot_triangle(): Ternary Plot for Three-Way Feature Relationships

This function visualizes three-part compositions (e.g., condition A/B/C contributions) on a ternary plot. Useful when analyzing data with three mutually exclusive categories or proportions summing to one.

This function is ideal for:

-   Displaying relationships between three mutually exclusive components.

-   Exploring feature allocation among three sources or pathways (e.g., tissue A/B/C).

-   Identifying samples/features located at edge or center of triangular composition space.


## Using a matrix or data.frame input

### Basic Usage

```{r}
samples <- c('pol1', 'pol2', 'pol3')
res_t = plot_triangle(
    x = df1,
    column_name = samples, 
    pch = 21,
    push_text = 1.3,
    col = c('#ff80e3', '#7eb1d5', '#fa8451')
)

title(main = 'Basic Usage', font.main = 1, cex = 0.5)
```

The points that locate at the center of the circle are those shared across three variables. As the points start moving towards the edges of the triangle they are dominated by one particular variable. Points that line over the perimeter of the triangle display shareness between the two variable those points are closer to.

The following aesthetics can be modified:

-   `background_color`

-   `cex`

-   `pch`

-   `col`

Now let's explore the features the plotting function offers.



### Example 1. Low Entropy Genes (0-0.4)

```{r}
res_t = plot_triangle(
    x = df1,
    column_name = samples, 
    pch = 21,
    push_text = 1.3,
    col = c('#ff80e3', '#7eb1d5', '#fa8451'),
    entropyrange = c(0, 0.4)
)

title(main = 'Low Entropy Genes', font.main = 1, cex = 0.5)
```

### Example 2. Medium Entropy Genes (0.4-1.3)

```{r}
res_t = plot_triangle(
    x = df1,
    column_name = samples, 
    pch = 21,
    push_text = 1.3,
    col = c('#ff80e3', '#7eb1d5', '#fa8451'),
    entropyrange = c(0.4, 1.3)
)

title(main = 'Medium Entropy Genes', font.main = 1, cex = 0.5)
```

### Example 3. High Entropy Genes (1.3-Inf)

```{r}
res_t = plot_triangle(
    x = df1,
    column_name = samples, 
    pch = 21,
    push_text = 1.3,
    col = c('#ff80e3', '#7eb1d5', '#fa8451'),
    entropyrange = c(1.3, Inf)
)

title(main = 'High Entropy Genes', font.main = 1, cex = 0.5)
```

### Example 4. High Entropy + Magnitude range (5-Inf)

```{r}
res_t = plot_triangle(
    x = df1,
    column_name = samples, 
    pch = 21,
    push_text = 1.3,
    col = c('#ff80e3', '#7eb1d5', '#fa8451'),
    entropyrange = c(1.3, Inf),
    maxvaluerange = c(5, Inf)
)

title(main = 'High Entropy Genes + Magnitude range', font.main = 1, cex = 0.5)
```

### Example 5. Remove Background Points

When setting `plotAll = FALSE` the function will remove all the points that fall outside of `entropyrange` and `maxvaluerange`

```{r}
res_t <- plot_triangle(
    x = df1,
    column_name = samples, 
    pch = 21,
    push_text = 1.3,
    col = c('#ff80e3', '#7eb1d5', '#fa8451'),
    entropyrange = c(1.3, Inf),
    maxvaluerange = c(5, Inf),
    plotAll = FALSE
)

title(main = 'Removing genes out of range', font.main = 1, cex = 0.5)
```

## Using a SummarizedExperiment object

The function can use any assay, the user should specify the name of this assay, if name is not specified by default it will look for the first assay in the se object

### Basic Usage

```{r}
samples = c('S1','S2','S3')
res_t <- plot_triangle(
    x = se,
   col = c('#7eb1d5','#ff80e3',  '#fa8451'),
    column_name = samples,
    assay_name = 'cpm_norm',
    pch = 21,
    push_text = 1.3,
)

title(main = 'Basic Usage SE', font.main = 1, cex = 0.5)
```

### Example 1. Low entropy (0-0.4)

```{r}
res_t <- plot_triangle(
    x = se,
   col = c('#7eb1d5','#ff80e3',  '#fa8451'),
    column_name = samples,
    assay_name = 'cpm_norm',
    entropyrange = c(0, 0.4),
    maxvaluerange = c(0.1, Inf),
    pch = 21,
    push_text = 1.3,
)

title(main = 'Low entropy SE', font.main = 1, cex = 0.5)
```

### Example 4. Output Data Retrieval

Data retrieval can be achieved by modifying the attribute `output_table` . The data contains information used for the plotting (This applies for any data structure used)

-   **`max_counts`** The maximum normalized expression value (across your selected samples) for that feature.

-   **`comx`**`comy`\
    The x– and y–coordinates used to place that point inside the triangle.

-   **`color`**\
    Which of your provided colors was applied (one per sample), or `whitesmoke` for filtered points.

-   **`entropy`**\
    The Shannon entropy score for that feature across all columns.

```{r}
res_t <- plot_triangle(
    x = se,
   col = c('#7eb1d5','#ff80e3',  '#fa8451'),
    column_name = samples,
    assay_name = 'cpm_norm',
    entropyrange = c(0, 0.4),
    maxvaluerange = c(0.1, Inf),
    pch = 21,
    push_text = 1.3, 
   output_table = TRUE
)

```

```{r}
# View first 6 rows of the output data
head(res_t)
```
## plot_circle: Entropy-Magnitude Circle Plot

This function visualizes high-dimensional input with any number of dimensions

Using a polar coordinate system where: - Radial position represents Shannon entropy degrees.

-   A total of P-1 (P = Number of Variables) circles are plotted and they represent degrees of dominance.

    -   The outtermost circle represents observations that are dominated only by one variable.

    -   The second outtermost circle represents observations that are dominated by two variables.

    -   The innermost circle represents observations that are uniform across all variables.

-   This function is ideal for:

<!-- -->

-   Visualizing multidimensional datasets (samples × features) in an interpretable 2D circular space.

-   Detecting samples/features with high entropy (irregularity) or high average expression.

-   Facilitating compact visualization across thousands of rows or columns.

**Basic Understanding:**

Dominant Sample: Shows which sample has the highest expression for each gene Useful for identifying sample-specific expression patterns

Radial Position: Genes near edge: Highly specific to one sample (low entropy) Genes near center: Similar expression across samples (high entropy)

Sector Position: Each wedge represents a sample Genes in a sample's wedge have their highest expression in that sample

```{r}
### Let's select only numerical columns
df2 = atac_tissue_score[,c(1,8:26)]
rownames(df2) = atac_tissue_score$Gene

```

The `circle_plot` function is able to handle SummarizedExperiment objects, data.frames and matrices just like the previous plots.

Note: there is some randomization in the asignment of the points in the space when a specific gene/observation is shared across many variables. In that way for example, if gene_abc is shared across tissue A, B, C and D, the function will randomly choose a variable for plotting 

## Using a dataframe

```{r}
plot_circle(
    x = df2,
    n = 19,
    entropyrange = c(0,Inf),
    magnituderange   = c(1, Inf),
    label  = 'legend',
    output_table = F,
    assay_name = 'cpm_norm'
)
```

## Using a SummarizedExperiment input


```{r}
plot_circle(
    x = se,
    n = 8,
    entropyrange = c(0,Inf),
    magnituderange   = c(0.1, Inf),
    label  = 'legend',
    output_table = FALSE,
    assay_name = 'cpm_norm'
)
```

The following aesthetics can be modified:

-   `background_polygon`

-   `background_na_polygon`

-   `background_alpha_polygon`

-   `point_size`

-   `point_fill_colors`

-   `point_fill_na_colors`

-   `point_line_colors`

-   `point_line_na_colors`

-   `straight_points`

-   `line_col`

-   `out_line`

-   `text_label_curve_size`

-   `label`

Now let's explore the features the plotting function offers using the our dataframe. The same plots can be made when using an `SummarizedExperiment`. Please see the examples for the `plot_circle` function `?plot_circle`

### Example 1: Low-entropy filtering (0-1)

Highlighting genes with low Entropy (high dominance in one variable)

```{r}
plot_circle(
    x = df2,
    n = 19,
    entropyrange     = c(0, 2),
    magnituderange   = c(1, Inf),
    label  = 'legend',
    output_table = FALSE
)
```

Highlighting samples is possible too. Lets highlight variable `Brain`, `Lung`, `Liver`. The names will appear around the circle if `label = 'curve'` and only those points and areas will be colored

```{r}
plot_circle(
    x = df2,
    n = 19,
    entropyrange     = c(0, 2),
    magnituderange   = c(1, Inf),
    label  = 'legend',
    variables_highlight = c('Brain', 'Lung', 'Liver', 'Breast'),
    output_table = FALSE,
    background_polygon = c('Brain' = '#7eb1d5', 'Lung' = 'gold', 
                           'Liver' = 'indianred', 'Breast' = '#ff80e3'), 
    point_fill_colors = c('Brain' = '#7eb1d5', 'Lung' = 'gold', 
                           'Liver' = 'indianred', 'Breast' = '#ff80e3'), 
    background_alpha_polygon = 0.1,
    point_size = 2, 
    point_line_colors = c('Brain' = 'black', 'Lung' = 'black', 
                          'Liver' = 'black', 'Breast' = 'black')
)


```

### Example 3: High-entropy filtering (2-3)

```{r}
plot_circle(
    x = df2,
    n = 19,
    entropyrange     = c(2, 3.5),
    magnituderange   = c(1, Inf),
    label  = 'legend',
    output_table = FALSE, 
    point_size = 3
)
```

### Example 4: Grouping by a column factor 'Category'

```{r}
plot_circle(
    x = df2,
    n = 19,
    entropyrange     = c(0, Inf),
    magnituderange   = c(1, Inf),
    column_variable_factor = 'core_type',
    label  = 'curve',
    output_table = FALSE, 
    point_size = 3
)
```

Let's highlight only a specific and spread the points

```{r}
plot_circle(
    x = df2,
    n = 19,
    entropyrange     = c(0, Inf),
    magnituderange   = c(1, Inf),
    column_variable_factor = 'core_type',
    label  = 'curve',
    output_table = FALSE, 
    point_size = 3,
    straight_points = FALSE, 
    point_fill_colors = c('Specific' = '#fa8451', '75% Tissues' = '#7eb1d5'),
    point_line_colors = c('Specific' = '#fa8451', '75% Tissues' = '#7eb1d5'), 
    
)

```

In this way it is possible to color observations by a specific category. 
In this example, we are highlighting RNA Pol III transcribed genes that have low uniformity
across tissues (aka = Specific) which locate at the edge of the circle. We are also highlighting
those genes with high uniformity across tissues (aka = 75% Tissues) which locate at the center of the circle

### Example 7: Retrieving plot data from SE

```{r}

results <- plot_circle(
    x = df2,
    n = 19,
    entropyrange     = c(0, Inf),
    magnituderange   = c(1, Inf),
    column_variable_factor = 'core_type',
    label  = 'curve',
    output_table = TRUE, 
    point_size = 3,
    straight_points = FALSE
    
)

```


The result is a list of two objects:

-   `results[[1]]`: a `ggplot2` object for visualization
-   `results[[2]]`: a `data.frame` with entropy, magnitude, etc.

```{r}
results[[1]]
head(results[[2]])

```

The returned data frame (`se_results[[2]]`) contains the following columns:

-   **Entropy**: the entropy score computed across rows (for each sample or feature).

-   **col**: the dominant sample alocated for that gene in the plotting space. When an observatio is tied among variables, a random variable is chosen

-   **rad**: the magnitude (mean expression) encoded as the radial distance in the plot.

-   **deg**: the plotting angle (in radians) for each sample’s axis.

-   **x**, **y**: the Cartesian coordinates corresponding to `(rad, deg)`, used internally by `geom_point()`.

-   **labels**: the text labels (e.g. sample names) when `label = "legend"` or `variables_highlight` is set.

-   **rand_deg**: the random rotation offset (fixed if you call `set.seed()` beforehand).

-   **alpha**: the point transparency (`1` for highlighted points, otherwise equal to your `background_alpha_polygon` setting).

## plot_circle_frequency(): Frequency-Stratified Entropy-Magnitude Visualization

This function builds upon plot_circle() by stratifying samples into frequency bins and visualizing entropy-magnitude patterns for each bin separately. Useful when your dataset contains variables/features with different levels of occurrence or sparsity (e.g., expressed vs. non-expressed genes).

This function is ideal for:

-   Identifying highly prevalent genes/features across a cohort.

-   Screening for outlier or inactive variables.

-   Visually comparing distributions in a compact format.

## Using a SummarizedExperiment object

```{r}
# Creating the circle plot data

# First we create the circle plot with output_table = TRUE to get 
# the data needed for the frequency plot. We'll use gene biotype as our 
# factor variable.

circle_data <- plot_circle(
    x = se,
    n = 8,
    column_variable_factor = 'gene_biotype',
    entropyrange = c(0, Inf),
    magnituderange = c(0.1, Inf),
    label = 'legend',
    output_table = TRUE,
    assay_name = 'cpm_norm'
)

```

### Example 1: Default parameters (combined panel)

```{r}
freq_plot_default <- plot_circle_frequency(
    n = 8,
    circle = circle_data,
    single = TRUE,
    legend = TRUE,
    numb_columns = 1,
    filter_class = NULL,
    point_size = 2
)

# Display the plot
freq_plot_default[[1]]

# View aggregated data
head(freq_plot_default[[2]])
```

### Example 2: Faceted by factor

```{r}
# Visualize each factor level in separate panels

plot_circle_frequency(
    n = 8,
    circle = circle_data,
    single = FALSE,
    legend = TRUE,
    numb_columns = 3,  # Arrange in 3 columns
    filter_class = NULL,
    point_size = 2
)
```

### Example 3: Filtering specific classes

```{r}
# Focus on specific gene biotypes

plot_circle_frequency(
    n = 8,
    circle = circle_data,
    single = FALSE,
    legend = TRUE,
    numb_columns = 1,  
    filter_class = c('protein_coding', 'snoRNA'),
    point_size = 3  
)
```

### Example 4: Combined plot with custom filtering

```{r}
# Create a combined plot showing only selected classes

plot_circle_frequency(
    n = 8,
    circle = circle_data,
    single = TRUE,
    legend = TRUE,
    numb_columns = 1,
     filter_class = c('protein_coding', 'snoRNA'),
    point_size = 3
)
```

### Output interpretation

Each arc segment represents:

-   A variable (e.g., gene), sorted by frequency.

-   Arc height indicates proportion of samples above threshold.

The returned table includes:

-   Variable: variable name (e.g., gene ID)

-   Proportion: % of samples with value above threshold

-   Threshold: cutoff used

-   Rank: position in sorted list

## Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```




